---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tvRRR

## Time-varying reduced rank regression using state-space models

<!-- badges: start -->
<!-- badges: end -->

**WORK IN PROGRESS**

The tvRRR package implements time-varying reduced rank regression as proposed in *A state-space approach to time-varying reduced-rank regression* (Work in progress, Brune, Bura and Scherrer, 2021+).  

## Installation

The development version can be installed from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("b-brune/tvRRR")
```

## Model

The model that is fitted is a reduced-rank regression with time-varying coefficient matrices. We implement two types of time-variation, called model (A) and model (B):

\[(A) \quad y_t = \alpha_t\beta'x_t + \Gamma{u}_t + \varepsilon_t\]

\[(B) \quad y_t = \alpha\beta_t'x_t + \Gamma{u}_t + \varepsilon_t\]

Model fitting is carried out using a Gaussian linear state-space model and the Kalman filter.

## Example application

The function `dataset()` draws data from models (A) and (B) with different specifications of the time-variation in the parameter matrix. 

```{r example}
library(tvRRR)

set.seed(712)
dat <- dataset("VARbreak", p = 5, d = 2, q = 5, t = 100, model = "A")
X <- dat$X; y <- dat$y
```

The main function of the dataset is `tvRRR()`. It automatically selects the rank of the model using BIC (in case it is unknown) and fits the tvRRR model.

```{r model fitting}
fit <- tvRRR(X = X, y = y, select_rank = TRUE, d_max = 5, silent = TRUE, 
             model = "A")

fit
```

`tvRRR()` returns a list of the filtered time-varying parameters (one-step-ahead, filtered and smoothed), as well as the fitted parameters and some information on the data log-likelihood and convergence. 

For more information on the model fitting algorithm see the paper, the accompanying vignette, and the function's documentation.
